Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-01-13 paper/paper-MinD.tex
start draft for next version of paper for PLOS computational biology Tested everything on bennet.

A number of models of the MinD protein system have been developed,
which accurately describe the dynamics using simulations of
reaction-diffusion equations.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.  A number of studies have used an approach similar in
that they do not rely on the ability of MinD to move along the walls
and cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically simulate the same
reaction-diffusion equations~\cite{fange2006noise, kerr2006division}.
These studies largely confirm the results of Huang's deterministic
model when applied to the wild-type, pill-shaped phenotype.  However,
the stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's 2003 model~\cite{huang2003dynamic} in
both deterministic and stochastic variants to study flattened cells
similar to those observed by Mannik \emph{et
  al.}~\cite{mannik2009bacterial}.

@@

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  In fact, when viewed as a movie
(see supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

We find agreement between our stochastic simulations and experiments
showing disrupted bipolar MinD oscillation in asymmetric flattened
E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations.  This result builds upon
existing results showing that this model is effective in a variety of
wild-type and mutated cell shapes~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, and reinforces that the stochastic
variation of Huang's 2003 model~\cite{fange2006noise,
  kerr2006division} has considerable predictive power beyond the
standard wild-type cell.

>>>>

\fixme{THE FOLLOWING NEEDS much CHANGING.  Many reaction-diffusion
  models have been proposed... There are several newer models that
  incorporate new experimental results.}
%
A number of models of the MinD protein system have been developed,
which accurately describe the dynamics using simulations of
reaction-diffusion equations.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.  A number of studies have used an approach similar in
that they do not rely on the ability of MinD to move along the walls
and cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically simulate the same \fixme{similar}
reaction-diffusion equations~\cite{fange2006noise, kerr2006division}.
These studies largely confirm the results of Huang's deterministic
model when applied to the wild-type, pill-shaped phenotype.  However,
the stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's 2003 model~\cite{huang2003dynamic} in
both deterministic and stochastic variants to study flattened cells
similar to those observed by Mannik \emph{et
  al.}~\cite{mannik2009bacterial}.  \fixme{Some weak
  justification... it has been widely used.}

@@

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  In fact, when viewed as a movie
(see supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  \fixme{Talk about
  membrane diffusion blurring out the starry night.}  This naturally
leads to irregularity in the temporal periodicity of the system, which
we quantify by examining the temporal correlation function of the
total MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

We find agreement between our stochastic simulations and experiments
showing disrupted bipolar MinD oscillation in asymmetric flattened
E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations.  This result builds upon
existing results showing that this model is effective in a variety of
wild-type and mutated cell shapes~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, and reinforces that the stochastic
variation of Huang's 2003 model~\cite{fange2006noise,
  kerr2006division} has considerable predictive power beyond the
standard wild-type cell.  \fixme{Discuss which of these are likely to
  map to more recent models.}
